miniKanren with fair search strategies

### 2019-04-28 Paper/paper.tex
fair

\newcommand{\conde}{\texttt{cond$^e$} }

@@

miniKanren programs, especially relational interpreters, have been proven to be useful in solving many problems \citep{byrd2017unified}. A subtlety in writing relational programs involving a large \conde expression, such as interpreters, is that the order of \conde clauses can affect the speed considerably. This is due to the biased allocation of computational resource.
Interleaving DFS, the search strategy of \conde in miniKanren~\citep{friedman_reasoned_2018}, allocates half resource to the first \conde clause, a quarter to the second, an eighth to the third, and so on. Under the hood, it is the unfairness of \texttt{disj}.

We address the unfairness of \texttt{disj} in this work. We propose two new search strategies, balanced interleaving DFS (biDFS) and breadth-first search (BFS). biDFS is almost fair -- the maximal ratio of resource among disjunctive goals is bounded by a constant factor. BFS is completely fair. We prove that our BFS is equivalent to the BFS proposed by Seres et al~\citep{seres1999algebra}. We also observe how new search strategies affect the efficiency and the answer order of known miniKanren programs.

@@

A search strategy is \emph{fair} if answers of lower costs always come first. 
The \emph{cost} of an answer is the number of relation applications. Now we 
illustrate the costs of answers by running a miniKanren relation. 
Fig.~\ref{def-repeato} defines a relation \texttt{repeato} which relates a term 
\texttt{x} with a list whose elements are all \texttt{x}s.

%TODO compare our definition with Silvija's, Oleg's, and Edward's.

\begin{figure}
  \lstinputlisting{Figures/defrel-repeato.rkt}
  \caption{\texttt{repeato}}
  \label{def-repeato}
\end{figure}

Consider the following \texttt{run} of \texttt{repeato}.
\begin{center}
  \begin{tabular}{c}
  \lstinputlisting{Figures/run-repeato-1.rkt}
   \end{tabular}
\end{center}

The above \texttt{run} generates 4 answers. All are lists of \texttt{*}s.
The order of the answers reflects the order miniKanren discovers them:
the first answer in the list is first discovered. This order is not surprising:
to generate the first answer, \texttt{'()}, miniKanren needs to apply
\texttt{repeato} only once and the later answers need more relation
applications. In this example, the cost of each answer is the same as
one more than the number of \texttt{*}s: the cost of \texttt{'()} is 1,
the cost of \texttt{'(*)} is 2, and so on.

% A query result (list of answers) is in the \emph{cost-respecting} order if no 
% answer occurs before another answer of a lower cost. In the above example, the 
% result is cost-respecting. The iDFS search, however, does not generate 
% cost-respecting answers in general. As an example, consider the following 
% \texttt{run} of \texttt{repeato}.

In the above example, every search strategy looks fair. However, the following 
example exposes that iDFS is not fair.

\begin{center}
  \begin{tabular}{c}
   \lstinputlisting{Figures/run-conde-repeato-iDFS.rkt}
   \end{tabular}
\end{center}

With iDFS, \texttt{'(a a)} occurs before \texttt{'(b)} while \texttt{'(a a)} is 
associated with a higher cost. iDFS strategy is the cause since it prioritizes 
the first \conde case considerably. When every \conde case 
are equally productive, the iDFS strategy takes $1/2^{i}$ answers from the 
$i$-th case, except the last case, which share the same portion as the second 
last one. In contrast, the same run with BFS produces answers in an expected 
order.

\begin{center}
  \begin{tabular}{c}
   \lstinputlisting{Figures/run-conde-repeato-BFS.rkt}
   \end{tabular}
\end{center}

Running the same query with biDFS results in yet another answer list. biDFS 
essentially organize disjunctive goals into a balanced tree. There is no way to 
build a balanced and complete tree of size 3, so one clause is allocated more 
resource than the other two.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/run-conde-repeato-biDFS.rkt}
	\end{tabular} 
\end{center}

If one insert a \texttt{(nevero)} as the forth clause, this run would results 
in the same answer list as the run with BFS. However, just making every \conde 
has $2^n$ clauses cannot turn biDFS to BFS.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/run-conde-repeato-biDFS-1.rkt}
	\end{tabular}
\end{center}

@@

>>>>

\newcommand{\conde}{\texttt{cond$^e$} }
\newcommand{\conj}{\texttt{conj}}
\newcommand{\disj}{\texttt{disj}}

@@

miniKanren programs, especially relational interpreters, have been proven to be useful in solving many problems \citep{byrd2017unified}. A subtlety in writing relational programs involving a large \conde expression, such as interpreters, is that the order of \conde clauses can affect the speed considerably.

Under the hood, \conde uses \texttt{conj} to combine goals within a clause, and \texttt{disj} to combine clauses. The \texttt{disj} in the current search strategy, interleaving DFS (iDFS), is unfair. It allocates half resource to its first goal, then allocates the other half to the rest similarly, except for the last clause that receives all the resource. This kind of \texttt{disj} prioritizes left clauses considerably in large \conde expressions.

Being aware of \texttt{disj} fairness, we also investigate \texttt{conj} fairness.

We propose three new search strategies, balanced interleaving DFS (biDFS), which has almost fair \disj, fair DFS (fDFS), which has fair \disj, and breadth-first search (BFS), which has fair \disj and fair \conj. The almost fair \texttt{disj} (i.e. biDFS) doesn't always allocate resource evenly, however, the maximal ratio of resource is bounded by 2. Fair\texttt{disj} allocates resource evenly to goals in the same disjunct. BFS has both fair \texttt{disj} and fair \texttt{conj}, where answers are generated in increasing order of cost. We prove that our BFS is equivalent to the BFS proposed by Seres et al~\citep{seres1999algebra}. We also observe how new search strategies affect the efficiency and answer order of known miniKanren programs.

@@

We illustrate how a fair search strategy should behavior by running a miniKanren relation, \texttt{repeato} (Fig.~\ref{repeato}), which relates a term \texttt{x} with a non-empty list whose elements are all \texttt{x}s.

\begin{figure}
	\lstinputlisting{Figures/repeato.rkt}
	\caption{\texttt{repeato} and an example run}
	\label{repeato}
\end{figure}

\subsection{fair \texttt{disj}}

In the following program, the three \conde clauses differs in trivial way. So we expect a search strategy with \emph{fair} \texttt{disj} would mix answers from each clause by grouping similar ones together. A search strategy is still considered fair if it permutes some lines of the answer list.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-fair-disj.rkt}
	\end{tabular}
\end{center}

\texttt{disj} in iDFS is not fair. Each clause takes half of the current resource and passes the other half to its following clauses, except for the last clause that takes all the current resource. Running the same query results in an answer list with about half of it being lists of \texttt{a}.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-iDFS.rkt}
	\end{tabular}
\end{center}

Our biDFS is almost fair. It gives a similar result in this case, except that the \texttt{b} clause takes most resource.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-biDFS-1.rkt}
	\end{tabular}
\end{center}

When the number of goals is a power of 2, biDFS would be fair.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-biDFS-2.rkt}
	\end{tabular}
\end{center}

\subsection{fair \texttt{conj}}

In the following program, the three \conde clauses differs in trivial way. So we expect a search strategy with \emph{fair} \texttt{conj} would mix answers derived from each clause by grouping similar ones together. A search strategy is still considered fair if it permutes some lines of the answer list. Strategies with unfair \conj produces more answers from the \texttt{a} clause. The result with biDFS is similar, except that \texttt{b} and \texttt{c} are swapped.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-conj-1.rkt}
	\end{tabular}
\end{center}

The program above has finite intermediate states (associating \texttt{x} to \texttt{'a}, \texttt{'b}, or \texttt{'c} respectively). Therefore, the result is a ``fair'' mix (in the sense of fair \texttt{disj}) of the three search spaces derived from each intermediate state. However, this approach is not applicable when the number of intermediate states is infinite. A solution also taken in \citep{seres1999algebra} is to organize answers as a stream of list, where each list has finite answer. And a \emph{fair} \texttt{conj} should allocate resource evenly among search spaces derived from states in the same bag.

A natural way to bag answers is by their costs. The \emph{cost} of an answer is the number of relation applications required to find them (their depth in search tree). In the previous examples, the costs of each answer is equal to their lengths. In the following example, the cost is equal to the length of inner list plus the length of outer list.

\begin{center}
	\begin{tabular}{c}
		\lstinputlisting{Figures/repeato-conj-2.rkt}
	\end{tabular}
\end{center}

% \section{fairness (old)}
% 
% A search strategy is \emph{fair} if answers of lower costs always come first. 
% The \emph{cost} of an answer is the number of relation applications. Now we 
% illustrate the costs of answers by running a miniKanren relation. 
% Fig.~\ref{repeato} defines a relation \texttt{repeato} which relates a term 
% \texttt{x} with a list whose elements are all \texttt{x}s.
% 
% %TODO compare our definition with Silvija's, Oleg's, and Edward's.
% 
% \begin{figure}
%   \lstinputlisting{Figures/defrel-repeato.rkt}
%   \caption{\texttt{repeato}}
%   \label{def-repeato}
% \end{figure}
% 
% Consider the following \texttt{run} of \texttt{repeato}.
% \begin{center}
%   \begin{tabular}{c}
%   \lstinputlisting{Figures/run-repeato-1.rkt}
%    \end{tabular}
% \end{center}
% 
% The above \texttt{run} generates 4 answers. All are lists of \texttt{*}s.
% The order of the answers reflects the order miniKanren discovers them:
% the first answer in the list is first discovered. This order is not surprising:
% to generate the first answer, \texttt{'()}, miniKanren needs to apply
% \texttt{repeato} only once and the later answers need more relation
% applications. In this example, the cost of each answer is the same as
% one more than the number of \texttt{*}s: the cost of \texttt{'()} is 1,
% the cost of \texttt{'(*)} is 2, and so on.
% 
% % A query result (list of answers) is in the \emph{cost-respecting} order if no 
% % answer occurs before another answer of a lower cost. In the above example, the 
% % result is cost-respecting. The iDFS search, however, does not generate 
% % cost-respecting answers in general. As an example, consider the following 
% % \texttt{run} of \texttt{repeato}.
% 
% In the above example, every search strategy looks fair. However, the following 
% example exposes that iDFS is not fair.
% 
% \begin{center}
%   \begin{tabular}{c}
%    \lstinputlisting{Figures/run-conde-repeato-iDFS.rkt}
%    \end{tabular}
% \end{center}
% 
% With iDFS, \texttt{'(a a)} occurs before \texttt{'(b)} while \texttt{'(a a)} is 
% associated with a higher cost. iDFS strategy is the cause since it prioritizes 
% the first \conde case considerably. When every \conde case 
% are equally productive, the iDFS strategy takes $1/2^{i}$ answers from the 
% $i$-th case, except the last case, which share the same portion as the second 
% last one. In contrast, the same run with BFS produces answers in an expected 
% order.
% 
% \begin{center}
%   \begin{tabular}{c}
%    \lstinputlisting{Figures/run-conde-repeato-BFS.rkt}
%    \end{tabular}
% \end{center}
% 
% Running the same query with biDFS results in yet another answer list. biDFS 
% essentially organize disjunctive goals into a balanced tree. There is no way to 
% build a balanced and complete tree of size 3, so one clause is allocated more 
% resource than the other two.
% 
% \begin{center}
% 	\begin{tabular}{c}
% 		\lstinputlisting{Figures/run-conde-repeato-biDFS.rkt}
% 	\end{tabular} 
% \end{center}
% 
% If one insert a \texttt{(nevero)} as the forth clause, this run would results 
% in the same answer list as the run with BFS. However, just making every \conde 
% has $2^n$ clauses cannot turn biDFS to BFS.
% 
% \begin{center}
% 	\begin{tabular}{c}
% 		\lstinputlisting{Figures/run-conde-repeato-biDFS-1.rkt}
% 	\end{tabular}
% \end{center}

@@

\section{fair DFS}
